Type-Safe Modular Parsing

### 2017-01-13 paper/sections/CaseStudy.tex
Mention some future work in conclusion.
Done 6.1

To demonstrate the utility of our modular parsing approach, we
implemented parsers of the first 18 calculi from book \textit{Types
  and Programming Languages} (TAPL) \cite{pierce2002types}. TAPL introduces several
calculi from simple to complex, by gradually adding new features to
syntax. We compared our implementation with a non-modular implementation
we found online, which is also written in Scala and uses the same Packrat parsing library.
We counted source lines of code (SLOC) and measured execution time for both implementations.
Our implementation saves almost 70\% code comparing with that non-modular one.

\huang{done}\bruno{mention that you compare against an existing non-modular implementation.}

\subsection{Implementation}

The calculi in the TAPL book are suitable for our case study for mainly two reasons.
Firstly, the calculi are arguably capture many of the language features
required in realistic programming languages. Secondly, the evolution of
calculi in the book reveals the advantages of modular representation
of abstract syntax and modular parsing, which is the key functionality
of our approach. By extracting common components from those calculi
and reusing them, we obtain considerably code reuse with reasonable performance penalty, as shown later.

\paragraph{Extracting Language Components}
Using the modular parsing techniques discussed before, we extract
reusable components from syntax of all the calculi. Each
component, which may contain several syntactical structures,
represents a certain feature of the language. For
example, the \lstinline{VarApp} component below represents variables and
function applications.

\huang{todo: move the code, rename PackratParser}
\begin{lstlisting}
object VarApp {
  trait Alg[E] {
    def TmVar(x: String): E
    def TmApp(e1: E, e2: E): E
  }
  trait Print extends Alg[String] {
    def TmVar(x: String): String = x
    def TmApp(e1: String, e2: String): String = "[" + e1 + " " + e2 + "]"
  }
  trait Parser[E] extends EParser[E] {
    lexical.delimiters += ("(", ")")
    val alg: Alg[E]
    val pVarAppE: PackratParser[E] = {
      lcid ^^ alg.TmVar |||
        pE ~ pE ^^ { case e1 ~ e2 => alg.TmApp(e1, e2) } |||
        "(" ~> pE <~ ")"
    }
  }
}
\end{lstlisting}

In our case study, each component is represented by a Scala object which has three parts.
\lstinline{Alg} is an object algebra interface for the abstract syntax.
\lstinline{Print} is an object algebra implements that interface
\lstinline{Alg}, for the pretty printing operation. They are standard object algebra operations.
\lstinline{Parser} is the modular parser for this piece of syntax.

We have some naming conventions in our code, \lstinline{E} represents
expressions, \lstinline{T} represents types and \lstinline{K}
represents kinds. They are the three sorts of syntax in our case study.
In this particular component \lstinline{VarApp} we only have expressions.

Similar as before, each parser carries lexical information, an object algebra field, and
parsing functions. We have some helper traits for convenience, such as \inlinecode{EParser}
containing \inlinecode{pE} for parsing expressions.

\begin{lstlisting}
trait EParser[E] {
  val pE: PackratParser[E]
}
\end{lstlisting}

@@

\begin{lstlisting}
object UntypedAbs {
  trait Alg[E] {
    def TmAbs(x: String, e: E): E
  }
  trait Print extends Alg[String] {
    def TmAbs(x: String, e: String): String = "\\" + x + "." + e
  }
  trait Parser[E] extends EParser[E] {
    lexical.delimiters += ("\\", ".", "(", ")")
    val alg: Alg[E]
    val pUntypedAbsE: PackratParser[E] =
      ("\\" ~> lcid) ~ ("." ~> pE) ^^ { case x ~ e0 => alg.TmAbs(x, e0) } |||
      "(" ~> pE <~ ")"
  }
}

object Untyped {
  trait Alg[E] extends UntypedAbs.Alg[E] with VarApp.Alg[E]
  trait Print extends Alg[String] with UntypedAbs.Print with VarApp.Print
  trait Parser[E] extends UntypedAbs.Parser[E] with VarApp.Parser[E] {
    override val alg: Alg[E]
    val pUntypedE: PackratParser[E] = pUntypedAbsE ||| pVarAppE
    override val pE: PackratParser[E] = pUntypedE
  }
}
\end{lstlisting}

When composing the calculus from components, all the traits can be
easily combined by Scala's \inlinecode{extends...with} keywords.
In \lstinline{Parser}, we need to override the object algebra interface and
parsing functions accordingly.

@@

\begin{lstlisting}
object TestUntyped {
  def work[E](inp: String)(a: Untyped.Alg[E]): E = {
    val lang = new Untyped.Parser[E] {
      override val alg: Untyped.Alg[E] = a
    }
    parse(lang.pE)(inp)
  }
  def parseAndPrint(inp: String): Unit = println(work(inp)(new Untyped.Print {}))
  def main(args: Array[String]): Unit = {
    parseAndPrint("(\\x.x) (\\y.y)")
  }
}
\end{lstlisting}

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{resources/depGraph.pdf}
    \caption{Dependency graph of all calculi and components}
    \label{fig:dependency}
\end{figure}

@@

Using our modular parsing approach, it is convenient to extract language
features as components and reuse them. From the dependency graph, we know
that common components such as \lstinline{VarApp} are reused in
lots of calculi. Such reuse could shorten the code considerably. We
will show this advantage and exam the possible performance penalty in
a quantitative way in the next subsection.

>>>>

To demonstrate the utility of our modular parsing approach, we
implemented parsers of the first 18 calculi from book \textit{Types
  and Programming Languages} (TAPL) \cite{pierce2002types}. TAPL introduces several
calculi from simple to complex, by gradually adding new features to
syntax. We compared our implementation with a non-modular implementation
we found online, which is also written in Scala and uses the same Packrat parsing library.
We counted source lines of code (SLOC) and measured execution time for both implementations.
Our implementation saves 69\% code comparing with that non-modular one.

\subsection{Implementation}\label{subsec:implementation}

The calculi in the TAPL book are suitable for our case study for mainly two reasons.
Firstly, they capture many of the language features
required in realistic programming languages, such as lambdas, records and variants. Secondly, the evolution of
calculi in the book reveals the advantages of modular representation
of abstract syntax and modular parsing, which is the key functionality
of our approach. By extracting common components from those calculi
and reusing them, we obtain considerably code reuse as shown later.

\paragraph{Extracting Language Components}
Using the pattern demonstrated in Section~\ref{subsec:language-component}, we extract
reusable components from all the calculi. Each
component, which may contain several syntactical structures,
represents a certain feature of the language. For
example, the \lstinline{VarApp} component below represents variables and
function applications, shown in Figure~\ref{fig:casestudy-varapp}.

\begin{figure}[ht]
\lstinputlisting[linerange=15-33]{../Scala/Parser/src/PaperCode/Sec6CaseStudy/Code1.scala}% APPLY:linerange=CASESTUDY_VARAPP
\caption{The \inlinecode{VarApp} component in case study.}\label{fig:casestudy-varapp}
\end{figure}

Similar as before, each component is represented by a Scala object which includes \lstinline{Alg} for the abstract syntax, \lstinline{Print} for pretty-printing, and \lstinline{Parse} for parsing.

We have some naming conventions in our code, \lstinline{E} represents
expressions, \lstinline{T} represents types and \lstinline{K}
represents kinds. They are the three sorts of syntax in our case study.
In the component \lstinline{VarApp} we only have expressions. We use some helper traits for eliminating duplicate definitions, such as \inlinecode{EParser} containing \inlinecode{pE} for parsing expressions.

\lstinputlisting[linerange=9-11]{../Scala/Parser/src/PaperCode/Sec6CaseStudy/Code1.scala}% APPLY:linerange=CASESTUDY_EPARSER

@@

The code of building \lstinline{Untyped} is presented in Figure~\ref{fig:casestudy-untyped}. Note that in the parser \inlinecode{Parse}, we need to override the object algebra interface and
parsing functions accordingly.

\begin{figure}[ht]
\lstinputlisting[linerange=37-64]{../Scala/Parser/src/PaperCode/Sec6CaseStudy/Code1.scala}% APPLY:linerange=CASESTUDY_UNTYPED
\caption{Build the \inlinecode{Untyped} calculus by composing to language components.}\label{fig:casestudy-untyped}
\end{figure}

@@

\lstinputlisting[linerange=68-78]{../Scala/Parser/src/PaperCode/Sec6CaseStudy/Code1.scala}% APPLY:linerange=CASESTUDY_DEMO

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{resources/depGraph.pdf}
    \caption{Dependency graph of all calculi and components in case study.}
    \label{fig:dependency}
\end{figure}

@@

From the dependency graph, we know
that common components such as \lstinline{VarApp} are reused in
lots of calculi. Such reuse could shorten the code considerably. We
will show this advantage and exam the possible performance penalty in
a quantitative way in the next subsection.
